---
title: "2b_full_tree_map_data"
output: html_document
date: "2026-07-07"
---

### Making a full dataset of street trees with metrics for all trees & whether or not they were damaged

#### 1. Make dataset of non-damaged trees

```{r}
# make dataset of non-serviced trees by filtering out trees in the serviced list
nonserviced_trees <- tree_map[!tree_map$tree_id %in% serviced_trees$tree_id, ]

# sanity check: 681637 + 29436 = 711037, but there are a lot of repeated entries in the serviced_trees set (trees that were damaged more than once) so this makes sense
```

#### 2. Resolve species to prep for getting height/crown width

```{r}
# extract species names from the nonserviced list
nonserviced_trees_species <- as.data.frame(table(nonserviced_trees$species_scientific))
nonserviced_trees_species <- as.character(nonserviced_trees_species$Var1)
```

```{r}
# resolve names and get higher taxonomic levels using GBIF
nonserviced_trees_resolved <- name_backbone_checklist(nonserviced_trees_species)
```

```{r}
# filter to necessary columns
nonserviced_trees_resolved <- nonserviced_trees_resolved %>%
  select("verbatim_name", "order", "family", "genus")

nonserviced_trees_resolved <- nonserviced_trees_resolved %>%
  rename(species_scientific = "verbatim_name")
```

#### 3. Join resolved species names to full tree list in order to run calculations

```{r}
# match resolved species names to individual nonserviced trees
nonserviced_trees <- left_join(nonserviced_trees, nonserviced_trees_resolved, by = "species_scientific")
```

#### 4. Estimate heights w/ DBH

```{r}
# same calculation process as the serviced trees
nonserviced_trees_matched_height <- nonserviced_trees %>%
  left_join(height_species_eqs, by = "species_scientific") %>%
  left_join(height_genus_eqs, by = "genus", suffix = c("", "_genus")) %>%
  left_join(height_family_eqs, by = "family", suffix = c("", "_family")) %>%
  left_join(height_order_eqs, by = "order", suffix = c("", "_order")) %>%
  mutate(
    model = coalesce(model, model_genus, model_family, model_order),
    B0    = coalesce(B0, B0_genus, B0_family, B0_order),
    B1    = coalesce(B1, B1_genus, B1_family, B1_order),
    B2    = coalesce(B2, B2_genus, B2_family, B2_order)
  ) %>%
  select(-ends_with("_genus"), -ends_with("_family"), -ends_with("_order"))
```

```{r}
nonserviced_trees_with_height <- nonserviced_trees_matched_height %>%
  mutate(height = calc_height(stump_diameter, model, B0, B1, B2))

nonserviced_trees_with_height %>%
  filter(is.na(height)) %>%
  count(species_scientific, genus, family, order, sort = TRUE)
```

#### 5. Estimate crown diameter w/ DBH

```{r}
nonserviced_trees_matched_crown <- nonserviced_trees %>%
  left_join(crown_species_eqs, by = "species_scientific") %>%
  left_join(crown_genus_eqs, by = "genus", suffix = c("", "_genus")) %>%
  left_join(crown_family_eqs, by = "family", suffix = c("", "_family")) %>%
  left_join(crown_order_eqs, by = "order", suffix = c("", "_order")) %>%
  mutate(
    model = coalesce(model, model_genus, model_family, model_order),
    B0    = coalesce(B0, B0_genus, B0_family, B0_order),
    B1    = coalesce(B1, B1_genus, B1_family, B1_order),
    B2    = coalesce(B2, B2_genus, B2_family, B2_order)
  ) %>%
  select(-ends_with("_genus"), -ends_with("_family"), -ends_with("_order"))
```

```{r}
nonserviced_trees_with_crown <- nonserviced_trees_matched_crown %>%
  mutate(crown_diameter = calc_crown(stump_diameter, model, B0, B1, B2))

nonserviced_trees_with_crown %>%
  filter(is.na(crown_diameter)) %>%
  count(species_scientific, genus, family, order, sort = TRUE)
```

#### 6. Create combined list of nonserviced trees & metrics

```{r}
nonserviced_trees_with_height <- nonserviced_trees_with_height %>%
  select("tree_id", "species_common", "species_scientific", "stump_diameter", "height", "closest_address", "geometry")

nonserviced_trees_with_crown <- nonserviced_trees_with_crown %>%
  select("crown_diameter")

nonserviced_trees_metrics <- cbind(nonserviced_trees_with_height, nonserviced_trees_with_crown)

nonserviced_trees_metrics <- nonserviced_trees_metrics %>%
    select("tree_id", "species_common", "species_scientific", "stump_diameter", "height", "crown_diameter", "closest_address", "geometry")
```

```{r}
# define coordinates as lat/long
st_crs(nonserviced_trees_metrics) <- 4326  
```

#### 7. full list, including repeats –\> 0s and 1s

```{r}
# combine datasets
tree_map_metrics <- rbind(nonserviced_trees_metrics, serviced_trees_metrics)
```

```{r}
# add 0/1 for no damage / damage
tree_map_metrics <- tree_map_metrics %>%
  mutate(serviced = as.integer(tree_id %in% serviced_trees_metrics$tree_id))
```

#### 8. Clean/remove negative values

```{r}
# drop negative values
tree_map_metrics$height[tree_map_metrics$height < 0] <- NA
tree_map_metrics$crown_diameter[tree_map_metrics$crown_diameter < 0] <- NA
```

```{r}
# save files
write.csv(nonserviced_trees_metrics, here("01_data", "1e_summary", "nonserviced_trees_metrics.csv"))
write.csv(tree_map_metrics, here("01_data", "1e_summary", "tree_map_metrics.csv"))
```

#### 9. full list without repeats for box plots

```{r}
tree_map_metrics_clean <- tree_map_metrics %>%
  distinct(geometry, .keep_all = TRUE)
```

10. make box plots of height, dbh, and crown diameter

```{r}
alltreesdbh <- ggplot(tree_map_metrics_clean, aes(x = "", y = stump_diameter)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Distribution of DBH",
    x = "",
    y = "DBH (inches)"
  ) +
  theme_minimal()
```

```{r}
alltreesdbh
```

```{r}
ggplot_build(alltreesdbh)$data[[1]]
```

```{r}
servicedtreesdbh <- ggplot(serviced_trees_metrics, aes(x = "", y = stump_diameter)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Distribution of Serviced Trees DBH",
    x = "",
    y = "DBH (inches)"
  ) +
  theme_minimal()

servicedtreesdbh
```

```{r}
ggplot_build(servicedtreesdbh)$data[[1]]
```

```{r}
nonservicedtreesdbh <- ggplot(nonserviced_trees_metrics, aes(x = "", y = stump_diameter)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Distribution of Non-serviced Trees DBH",
    x = "",
    y = "DBH (inches)"
  ) +
  theme_minimal()

nonservicedtreesdbh
```

```{r}
ggplot_build(nonservicedtreesdbh)$data[[1]]
```

saving to plot in arcgis

```{r}
serviced_trees_gis <- serviced_trees %>%
  mutate(
    longitude = st_coordinates(.)[, "X"],
    latitude = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()
```

```{r}
write_csv(serviced_trees_gis, "serviced_trees_gis.csv")
```

```{r}
tree_map_gis <- tree_map %>%
  mutate(
    longitude = st_coordinates(.)[, "X"],
    latitude = st_coordinates(.)[, "Y"]
  ) %>%
  st_drop_geometry()
```

```{r}
write_csv(tree_map_gis, "tree_map_gis.csv")
```

Filtering down to approved trees

```{r}
approved_species <- read.csv("nyc_approved_species.csv")
```

```{r}
approved_genus_only <- approved_species$species_scientific[str_count(approved_species$species_scientific, "\\S+") == 1] 
approved_full_species <- approved_species$species_scientific[str_count(approved_species$species_scientific, "\\S+") > 1] 
```

```{r}
approved_all_genera <- unique(c(
  approved_genus_only,
  word(approved_full_species, 1)
))
```

```{r}
tree_map_genera <- tree_map_metrics %>%
  mutate(genus = word(species_scientific, 1))
```

```{r}
trees_approved <- tree_map_genera %>%
  filter(
    species_scientific %in% approved_full_species |                          # rule 1: exact species match
      genus %in% approved_genus_only |                                          # rule 2: whole genus approved
      (species_scientific == genus & genus %in% approved_all_genera)            # rule 3: genus-only record, genus represented on approved list
  )
```

```{r}
#filtering down to serviced tree species
serviced_trees_approved <- trees_approved %>%
  filter(trees_approved$serviced == "1")
```

```{r}
# adding NTA names to dataset
trees_approved <- st_as_sf(trees_approved)
trees_approved <- st_transform(trees_approved, st_crs(neighborhood_boundaries))

trees_approved <- st_join(trees_approved, neighborhood_boundaries, join = st_within)
```

```{r}
trees_approved <- trees_approved %>%
  select("tree_id", "species_common", "species_scientific", "genus", "stump_diameter", "height", "crown_diameter", "closest_address", "serviced", "ntaname", "geometry")
```

```{r}
write.csv(trees_approved, "trees_approved.csv", row.names = FALSE)
write.csv(serviced_trees_approved, "serviced_trees_approved.csv", row.names = FALSE)
```
