---
title: "2b_full_tree_map_data"
output: html_document
date: "2026-07-07"
---

### Making a full dataset of street trees with metrics for all trees & whether or not they were damaged

#### 1. Make dataset of non-damaged trees

```{r}
nonserviced_trees <- tree_map[!tree_map$tree_id %in% serviced_trees$tree_id, ]
# sanity check: 681637 + 29436 = 711037, but there are a lot of repeated entries in the serviced_trees set (trees that were damaged more than once) so this makes sense
```

#### 2. Resolve species to prep for getting height/crown width

```{r}
nonserviced_trees_species <- as.data.frame(table(nonserviced_trees$species_scientific))
nonserviced_trees_species <- as.character(nonserviced_trees_species$Var1)
```

```{r}
nonserviced_trees_resolved <- name_backbone_checklist(nonserviced_trees_species)
```

```{r}
nonserviced_trees_resolved <- nonserviced_trees_resolved %>%
  select("verbatim_name", "order", "family", "genus")

nonserviced_trees_resolved <- nonserviced_trees_resolved %>%
  rename(species_scientific = "verbatim_name")
```

now we have a dataframe listing all of the species and their resolved names + higher orders, next step is to join this to the nonserviced list so that each indiv tree has its higher classifications –\> do DBH estimates!
