---
title: "2b_neighborhood_service_requests"
output: html_document
date: "2026-07-01"
---

```{r}
file_move(path = "/Users/anndai/Downloads/2020_Neighborhood_Tabulation_Areas_(NTAs)_20260624.geojson", new_path = "/Users/anndai/street_tree_storm_resistance_2026")
```

```{r}
neighborhood_boundaries <- st_read("2020_Neighborhood_Tabulation_Areas_(NTAs)_20260624.geojson")
```

```{r}
names(neighborhood_boundaries)
head(neighborhood_boundaries)
```

```{r}
st_crs(tree_map)
st_crs(neighborhood_boundaries)

alltrees <- st_transform(tree_map, st_crs(neighborhood_boundaries))
damagedtrees <- st_transform(serviced_trees, st_crs(neighborhood_boundaries))
```

```{r}
trees_joined <- st_join(alltrees, neighborhood_boundaries, join = st_within)
damaged_joined <- st_join(damagedtrees, neighborhood_boundaries, join = st_within)
```

```{r}
total_counts <- trees_joined %>%
  st_drop_geometry() %>%
  count(ntaname, name = "total_trees")

damaged_counts <- damaged_joined %>%
  st_drop_geometry() %>%
  count(ntaname, name = "damaged_trees")

ratio_df <- total_counts %>%
  left_join(damaged_counts, by = "ntaname") %>%
  mutate(
    damaged_trees = ifelse(is.na(damaged_trees), 0, damaged_trees),
    damage_ratio = damaged_trees / total_trees
  ) %>%
  arrange(desc(damage_ratio))
```

```{r}
neighborhoods_ratio <- neighborhood_boundaries %>%
  left_join(ratio_df, by = "ntaname")

ggplot(neighborhoods_ratio) +
  geom_sf(aes(fill = damage_ratio), color = "white", linewidth = 0.1) +
  scale_fill_distiller(
     palette = "YlOrRd",
  direction = 1,
  trans = "sqrt",
  labels = scales::percent,
  na.value = "grey90",
    name = "Damage Ratio"
  ) +
  labs(title = "Storm Damage Ratio by NYC Neighborhood") +
  theme_void()
```

This is quite interesting - even after adjusting for the number of trees total in each neighborhood, the non-EJ Areas still have a higher damage ratio. I guess we can look at this from a different perspective: this is the ratio of service requests to total trees. Non-EJ Areas in general had higher numbers of service requests. This might be due to social factors: a higher number of non-English speakers, less access to phones/internet to send in requests in general, etc? What does this say about other types of 311 requests in severe weather situations?
