---
title: "introductory_tables" output: html_document ---
---

```{r}
library(ggplot2)
```

```{r}
#frequency table of all street trees in NYC
street_tree_counts <- tree_map %>%
  group_by(species_scientific) %>%
  summarise(frequency = n()) %>%
  arrange(desc(frequency))

#cut to top 15
top15_street_trees <- street_tree_counts %>%
  slice_max(frequency, n = 15)
```

```{r}
#cut to top 15
top30_street_trees <- street_tree_counts %>%
  slice_max(frequency, n = 30)
```

```{r}
#frequency table of serviced trees
serviced_tree_counts <- serviced_trees %>%
  group_by(species_scientific) %>%
  summarise(frequency = n()) %>%
  arrange(desc(frequency))

serviced_tree_counts$geometry <- NULL

#cut to top 15
top30_serviced_trees <- serviced_tree_counts %>%
  slice_max(frequency, n = 30)
```

```{r}
#bar graph of top 15 street tree species
top30_street_trees %>%
  ggplot(aes(x = reorder(species_scientific, frequency), y=frequency)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

```{r}
#bar graph of top 15 serviced trees
top30_serviced_trees %>%
  ggplot(aes(x = reorder(species_scientific, frequency), y=frequency)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

```{r}
nyc_approved_trees <- read.csv("nyc_approved_trees.csv")
```

```{r}
approved_tree_frequency <- semi_join(street_tree_counts, nyc_approved_trees, by = "species_scientific")
```

```{r}
approved_tree_frequency %>%
  ggplot(aes(x = reorder(species_scientific, frequency), y=frequency)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

```{r}
# Flag matches
top30_street_trees <- top30_street_trees %>%
  mutate(highlight = ifelse(species_scientific %in% nyc_approved_trees$species_scientific, 
                             "Match", "No Match"))

# Plot
ggplot(top30_street_trees, aes(x = reorder(species_scientific, frequency), y = frequency, fill = highlight)) +
  geom_bar(stat = "identity") +  coord_flip() +
  scale_fill_manual(values = c("Match" = "forestgreen", "No Match" = "grey75")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
top30_serviced_trees <- top30_street_trees %>%
  mutate(highlight = ifelse(species_scientific %in% nyc_approved_trees$species_scientific, 
                             "Match", "No Match"))

# Plot
ggplot(top30_serviced_trees, aes(x = reorder(species_scientific, frequency), y = frequency, fill = highlight)) +
  geom_bar(stat = "identity") +  coord_flip() +
  scale_fill_manual(values = c("Match" = "forestgreen", "No Match" = "grey75")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#service requests by borough
servicerequest_location <- serviced_trees %>%
  group_by(borough) %>%
  summarise(frequency = n()) %>%
  arrange(desc(frequency))
```

```{r}
#plotting it
#seems to be quite proportional to the number of trees total in each borough (queens: 4.3%, mh: 1.6%)
servicerequest_location %>%
  ggplot(aes(x = reorder(borough, frequency), y=frequency)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

-   heatmap of service requests by neighborhood

-   how to get trees distance from each other

-   once you get the species –\> use citree to see characteristics of top 10 bottom 10

```{r}
#merge rows that were counted twice
serviced_trees <- serviced_trees %>%
  distinct(geometry, service_request_date, .keep_all = TRUE)
```

```{r}
# tree species that were serviced more than once
repeated_requests <- serviced_trees[duplicated(serviced_trees$geometry) | duplicated(serviced_trees$geometry, fromLast = TRUE), ]

repeated_requests <- serviced_trees %>%
  group_by(geometry) %>%
  mutate(geom_count = n()) %>%
  filter(geom_count > 1) %>%
  arrange(geometry)

repeated_requests <- repeated_requests %>%
  distinct(storm_event_id, .keep_all = TRUE)
```

```{r}
repeated_requests <- repeated_requests %>%
  group_by(geometry) %>%
  filter(n() > 1) %>%
  ungroup()
```

```{r}
repeated_request_counts <- length(unique(repeated_requests$geometry))
# 1184 trees were damaged more than once
# ^ need to cut to take out trees that were called multiple times in the same storm (based on just date)
```

-   what proportion of the total population is each species

-   what proportion of the total service requests is each species

-   what proportion of the total species population are the service request trees (what you did before with the ratios)

```{r}
#ratio of each species to the total number of trees in NYC
total_tree_ratios <- street_tree_counts %>%
  mutate(
    total = sum(frequency),
    ratio = frequency / 703283, percentage = round(ratio * 100, 2)
  )
```

```{r}
# ratio of each serviced species count to the total number of serviced trees
serviced_tree_ratios <- serviced_tree_counts %>%
  mutate(
    total = sum(frequency),
    ratio = frequency / 29436, percentage = round(ratio * 100, 2)
  )
```

```{r}
# ratio of serviced trees for each population to total species population
serviced_species_ratios <- serviced_tree_counts %>% left_join (street_tree_counts, by = "species_scientific")

serviced_species_ratios <- serviced_species_ratios %>%
  mutate(ratio = frequency.x / frequency.y, 
percentage = round(ratio * 100, 2))

```

with genus now...?

```{r}
serviced_tree_genus_counts <- serviced_tree_counts %>%
  mutate(genus = word(species_scientific, 1)) %>%
  group_by(genus) %>%
  summarise(frequency = sum(frequency)) %>%
  arrange(desc(frequency))
```

```{r}
street_tree_genus_counts <- street_tree_counts %>%
  mutate(genus = word(species_scientific, 1)) %>%
  group_by(genus) %>%
  summarise(frequency = sum(frequency)) %>%
  arrange(desc(frequency))
```

```{r}
serviced_genus_ratios <- serviced_tree_genus_counts %>% left_join (street_tree_genus_counts, by = "genus")

serviced_genus_ratios <- serviced_genus_ratios %>%
  mutate(ratio = frequency.x / frequency.y, 
percentage = round(ratio * 100, 2))
```

```{r}
total_genus_ratios <- street_tree_genus_counts %>%
  mutate(
    total = sum(frequency),
    ratio = frequency / 703283, percentage = round(ratio * 100, 2)
  )
```

```{r}
serviced_genus_ratios <- serviced_tree_genus_counts %>%
  mutate(
    total = sum(frequency),
    ratio = frequency / 29436, percentage = round(ratio * 100, 2)
  )
```
