---
title: "2c_regression_model"
output: html_document
date: "2026-07-20"
---

Building a regression model in layers to test association between tree species/genus and physical damage from severe weather

1.  first-generation model: logistic regression using GLM, finding intercept, random effect, and slope for approved tree species and storm damage
    - species as fixed effect, location as random effect
    - confounders: DBH, height, crown diameter
    - reference species: Quercus rubra (median damage rate for comparison as a "typical" species)

```{r}
# determining reference species
trees_approved %>%
  group_by(species_scientific) %>%
  summarize(pct_damaged = mean(serviced, na.rm = TRUE), n = n()) %>%
  filter(n >= 100) %>%
  arrange(abs(pct_damaged - median(pct_damaged)))

# using Prunus serrulata, as its damage rate is close to the median but with a larger sample size compared to Quercus rubra etc
```

```{r}
# setting reference species
trees_approved$species_scientific <- relevel(factor(trees_approved$species_scientific), ref = "Prunus serrulata")
```

```{r}
install.packages("glmmTMB")
library(glmmTMB)

model_1 <- glmmTMB(serviced ~ species_scientific + stump_diameter + height + crown_diameter + (1 | ntaname),
               data = trees_approved,
               family = binomial)

summary(model_1)
```

```{r}
# model with just DBH as confounder
model_2 <- glmmTMB(serviced ~ species_scientific + stump_diameter + (1 | ntaname),
                   data = trees_approved,
                   family = binomial)

summary(model_2)
```

```{r}
# model without predictors
model_3 <- glmmTMB(serviced ~ species_scientific + (1 | ntaname),
                   data = trees_approved,
                   family = binomial)

summary(model_3)
```

```{r}
model_4 <- glmmTMB(serviced ~ species_scientific - 1 + (1 | ntaname),
                 data = trees_approved,
                 family = binomial)

summary(model_4)
```

```{r}
exp(fixef(model_3)$cond)
```
