---
title: "2c_regression_model"
output: html_document
date: "2026-07-20"
---

Building a regression model in layers to test association between tree species/genus and physical damage from severe weather

1.  first-generation model: logistic regression using GLM, finding intercept, random effect, and slope for approved tree species and storm damage
    - species as fixed effect, location as random effect
    - confounders: DBH, height, crown diameter
    - reference species: Quercus rubra (median damage rate for comparison as a "typical" species)

```{r}
# determining reference species
trees_approved %>%
  group_by(species_scientific) %>%
  summarize(pct_damaged = mean(serviced, na.rm = TRUE), n = n()) %>%
  filter(n >= 100) %>%
  arrange(abs(pct_damaged - median(pct_damaged)))

# using Prunus serrulata, as its damage rate is close to the median but with a larger sample size compared to Quercus rubra etc
```

```{r}
# setting reference species
trees_approved$species_scientific <- relevel(factor(trees_approved$species_scientific), ref = "Prunus serrulata")
```

```{r}
install.packages("glmmTMB")
```

```{r}
library(glmmTMB)
```

```{r}
install.packages("glmmTMB")
library(glmmTMB)

model_1 <- glmmTMB(serviced ~ species_scientific + stump_diameter + height + crown_diameter + (1 | ntaname),
               data = trees_approved,
               family = binomial)

summary(model_1)
```

```{r}
# model with just DBH as confounder
model_2 <- glmmTMB(serviced ~ species_scientific + stump_diameter + (1 | ntaname),
                   data = trees_approved,
                   family = binomial)

summary(model_2)
```

```{r}
# model without predictors
model_3 <- glmmTMB(serviced ~ species_scientific + (1 | ntaname),
                   data = trees_approved,
                   family = binomial)

summary(model_3)
```

```{r}
model_4 <- glmmTMB(serviced ~ species_scientific - 1 + (1 | ntaname),
                 data = trees_approved,
                 family = binomial)

summary(model_4)
```

```{r}
exp(fixef(model_3)$cond)
```

```{r}
library(broom.mixed)

plot_model_3 <- tidy(model_3, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%   # drop intercept if present
  mutate(
    term = gsub("species_scientific", "", term),  # clean up label text
    estimate = exp(estimate),          # convert log-odds -> odds ratio
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

plot_model_3_top_bottom <- bind_rows(
  plot_model_3 %>% slice_max(estimate, n = 15) %>% mutate(group = "Highest Risk"),
  plot_model_3 %>% slice_min(estimate, n = 15) %>% mutate(group = "Lowest Risk")
)

ggplot(plot_model_3_top_bottom, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  labs(
    title = "Storm Damage Risk by Species (Odds Ratios)",
    x = "Odds Ratio (log scale)",
    y = "",
    color = ""
  ) +
  theme_minimal()
```

```{r}
plot_model_3_filtered <- plot_model_3 %>%
  left_join(trees_approved %>% count(species_scientific), by = c("term" = "species_scientific")) %>%
  filter(n >= 100)
```

```{r}
plot_model_3_top_bottom <- bind_rows(
  plot_model_3_filtered %>% slice_max(estimate, n = 15) %>% mutate(group = "Highest Risk"),
  plot_model_3_filtered %>% slice_min(estimate, n = 15) %>% mutate(group = "Lowest Risk")
)

ggplot(plot_model_3_top_bottom, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  labs(
    title = "Storm Damage Risk by Species (Odds Ratios)",
    x = "Odds Ratio (log scale)",
    y = "",
    color = ""
  ) +
  theme_minimal()
```

```{r}
plot_model_2 <- tidy(model_2, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%   # drop intercept if present
  mutate(
    term = gsub("species_scientific", "", term),  # clean up label text
    estimate = exp(estimate),          # convert log-odds -> odds ratio
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

plot_model_2_filtered <- plot_model_2 %>%
  left_join(trees_approved %>% count(species_scientific), by = c("term" = "species_scientific")) %>%
  filter(n >= 100)

plot_model_2_top_bottom <- bind_rows(
  plot_model_2_filtered %>% slice_max(estimate, n = 15) %>% mutate(group = "Highest Risk"),
  plot_model_2_filtered %>% slice_min(estimate, n = 15) %>% mutate(group = "Lowest Risk")
)

ggplot(plot_model_2_top_bottom, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  labs(
    title = "Storm Damage Risk by Species (Odds Ratios)",
    x = "Odds Ratio (log scale)",
    y = "",
    color = ""
  ) +
  theme_minimal()
```
