---
title: "2b_stormdata_pca"
output: html_document
date: "2026-08-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Principal Component Analysis

Using ML, turning metrics for each storm event (from ERA5) into an intensity number that we can use to control for the effect of storm intensity on tree damage in the regression model.

#### 1. Get PCA1 severity score for each storm

```{r}
# upload storm data
storms_era5 <- read_csv(here("01_data", "1a_raw", "nyc_storm_hazards_era5_by_borough.csv"))
```

```{r}
# filter down to selected storm events --> just all NYC data for each storm event
storms_filtered <- storms_era5 %>% 
  filter(event_id %in% hazard_history$storm_event_id)
```

```{r}
# the era5 data includes borough-specific data, but here we filter those rows down to only the affected boroughs of each storm (as reported in boroughs_reported)

storms_byborough <- storms_filtered %>%
  rowwise() %>%
  filter(unit %in% str_split(boroughs_reported, ",\\s*")[[1]]) %>%
  ungroup()
```

```{r}
# load library for PCA function
library(stats)
```

```{r}
# run PCA
storm_metrics <- storms_byborough %>%
  select(wind_peak_ms, wind_mean_ms, wind_peak_cellmax_ms,
         precip_total_mm, precip_peak_rate_mmh, precip_peak_rate_cellmax_mmh)

pca_result <- prcomp(storm_metrics, scale. = TRUE, center = TRUE)

summary(pca_result)          # check variance explained by PC1
pca_result$rotation[, 1]     # check loadings/direction
```

```{r}
storms_byborough <- storms_byborough %>%
  mutate(
    severity_raw = pca_result$x[, 1],
    severity_1_10 = scales::rescale(severity_raw, to = c(1, 10))
  )
```

```{r}
storms_byborough <- storms_byborough %>%
  rename(borough = unit)
```

```{r}
storms_byborough <- storms_byborough %>%
  rename(storm_id = event_id)
```

```{r}
serviced_trees_metrics <- serviced_trees_metrics %>%
  rename(storm_id = storm_event_id)
```

```{r}
storms_byborough <- storms_byborough %>%
  rename(severity = severity_1_10)
```

#### 2. Join serviced trees with specific severity score

```{r}
serviced_trees_metrics <- serviced_trees_metrics %>%
  select(-borough) %>%   # drop the old, mismatched text column
  st_join(boroughs %>% select(boroname), join = st_within) %>%
  rename(borough = boroname)   # adjust to your actual boundary column name
```

```{r}
serviced_trees_full <- serviced_trees_metrics %>%
  st_drop_geometry() %>%
  left_join(storms_byborough %>% select(storm_id, borough, severity), by = c("storm_id", "borough"))

serviced_trees_full <- serviced_trees_full %>%
  mutate(serviced = 1)
```

```{r}
# get closest borough severity for trees w/o borough severity match
missing_severity <- serviced_trees_full %>%
  filter(!is.na(borough), is.na(severity))

nrow(missing_severity)
```

```{r}
library(purrr)

fill_nearest_borough_severity <- function(tree_row, storms_by_borough, boroughs) {
  storm <- tree_row$storm_id
  
  # boroughs that DO have severity data for this specific storm
  available_boroughs <- storms_byborough %>%
    filter(storm_id == storm, !is.na(severity)) %>%
    pull(borough)
  
  if (length(available_boroughs) == 0) return(NA_real_)  # no borough has data for this storm at all
  
  # subset the borough polygons to just those available
  boroughs_available <- boroughs %>% filter(boroname %in% available_boroughs)
  
  # find nearest among just those
  nearest_idx <- st_nearest_feature(tree_row, boroughs_available)
  nearest_borough <- boroughs_available$boroname[nearest_idx]
  
  # pull that borough's severity for this storm
  storms_by_borough %>%
    filter(storm_id == storm, borough == nearest_borough) %>%
    pull(severity) %>%
    first()
}
```

```{r}
missing_severity <- st_as_sf(missing_severity)
missing_severity <- st_transform(missing_severity, st_crs(boroughs))
```

```{r}
missing_severity$severity <- map_dbl(seq_len(nrow(missing_severity)), function(i) {
  fill_nearest_borough_severity(missing_severity[i, ], storms_byborough, boroughs)
})
```

```{r}
serviced_trees_full <- serviced_trees_full %>%
  filter(!(is.na(severity) & !is.na(borough))) %>%   # remove the rows we're about to replace
  bind_rows(missing_severity)

serviced_trees_full <- serviced_trees_full %>% drop_na(borough)
```

#### 3. Join nonserviced trees with average borough storm severity score

```{r}
borough_avg_severity <- storms_byborough %>%
  group_by(borough) %>%
  summarize(severity = mean(severity, na.rm = TRUE))

nonserviced_trees_full <- nonserviced_trees_metrics %>%
  left_join(borough_avg_severity, by = "borough") %>%
  mutate(serviced = 0)
```

#### 4. Join datasets together for full treemap data –\> regression

```{r}
serviced_trees_join <- serviced_trees_full %>%
  select()

nonserviced_trees_join <- nonserviced_trees_full %>%
  select()

rbind(serviced_trees_join, nonserviced_trees_join)
```
